---
title: "In Class Exercise 2: Working with Master Plan Planning Sub-zone Data"
author: "Jiale SO"
date: "August 26, 2024"
date-modified: "last-modified"
execute: 
  eval: true  
  echo: true
  freeze: true
  warning: false
  message: false
---

# Exercise 2 Working with Master Plan Planning Sub-zone data.

# 1.0 Introduction

# 2.0 Setup and Requirements

```{r}
pacman::p_load(sf, tmap, tidyverse, dplyr)
```

# 3.0 Importing the Data Sets.

## 3.1 Importing 2014 SHPZ File

::: panel-tabset
### Task at Hand

1.  Go to data.gov.sg and download the 2014 Master Subzone SHP and KML File
2.  Import and read the SHP and KML File in R.

### SHPZ File Importing

We can use the `st_read` function to import the data.

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

### KML File Importing

Similarly we can use the `st_read` function to import the data.

```{r, error=TRUE}
mspz_kml <- st_read("data/geospatial/MasterPlan2014SubzoneBoundaryWebKML.kml")
```

However, the data on data.gov.sg is corrupted. Hence we will do a workaround, where we can create a new kml file base on the shp file and read it.

Step 1: using the `st_write` to create a kml file from the shapefile

Step 2: using the `st_read` function to read the generated kml file.

```{r, error=TRUE}
st_write(mpsz, "data/geospatial/MP14_SUBZONE_WEB_PL.kml", driver = "KML")
mpsz_kml <- st_read("data/geospatial/MP14_SUBZONE_WEB_PL.kml")
```
:::

## 3.2 Importing 2019 SHPZ File

::: panel-tabset
## Task At Hand

1.  Go to data.gov.sg and download the 2019 KML File
2.  Import and read the KML File in R.

## Importing the SHP File

```{r}
mpsz19_shp <- st_read(dsn = "data/geospatial", layer = "MPSZ-2019")
```

## Importing KML File

```{r}
mpsz19_kml <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```
:::

## 3.3 Comparison the difference between SHP and KML File and formatting the CRS.

### 3.3.1 Understanding SHP and KML Files

Use `glimpse()` and `head()` to compare the attributes available in each file. For instance, SHP files might have more detailed attribute tables with multiple fields, while KML files are more streamlined with fewer attributes.

::: panel-tabset
## SHP File

```{r}
glimpse(mpsz19_shp)
```

```{r}
head(mpsz19_shp)
```

## KML FILE

```{r}
glimpse(mpsz19_kml)
```

```{r}
head(mpsz19_kml)
```
:::

### SHP vs. KML: Key Differences

### Summary Table:

| Property            | SHP (Shapefile)                              | KML (Keyhole Markup Language)     |
|---------------------|----------------------------------------------|-----------------------------------|
| **File Structure**  | Multiple files for geometries and attributes | Single file for visualization     |
| **Attributes**      | Detailed, diverse data types                 | Limited, mostly descriptive       |
| **Geometry Types**  | Complex, suited for GIS analysis             | Basic, suited for display         |
| **CRS Information** | Explicit and configurable (e.g., EPSG:3414)  | Typically defaults to WGS84       |
| **Usage Context**   | GIS analysis, mapping                        | Web-based visualization           |
| **Performance**     | Slower with large datasets                   | Lightweight and quick for display |

### 3.3.2 CRS (Coordinate Reference System)

-   **SHP File CRS:** Shapefiles often come with a defined CRS. In this case, we ensure that it is aligned with Singapore’s local system (EPSG:3414).

-   **KML File CRS:** KML files typically default to WGS84 (EPSG:4326), which is used in web applications and is latitude/longitude-based.

::: {.callout-tip appearance="simple"}
How to know which CRS to use ? We can simply go to epsg.io to check <https://epsg.io/?q=singapore>
:::

```{r}
st_crs(mpsz19_shp)
```

```{r}
mpsz19_shp_crs3414 <- st_read(dsn = "data/geospatial", layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
st_crs(mpsz19_shp_crs3414)
```

# 4.0 Working with Population Data

## 4.1 Import and Simple Data Transformation

::: panel-tabset
## The Task

Go to[SingStat Website](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) and download 2023's dataset for our popuplation data. We then group them by age and subzone

## Code

```{r}
popdata <- read_csv("data/respopagesextod2023.csv")
popdata2023 <- popdata %>%
  group_by(PA, SZ, AG) %>%  # Ensure data is grouped by 'AG' if needed
  summarise(POP = sum(Pop, na.rm = TRUE)) %>%  # Summarise population
  pivot_wider(names_from = AG, values_from = POP) %>%  # Pivot to wider format
  ungroup()  # Ungroup the data

# Check the column names
colnames(popdata2023)

```
:::

## 4.2 Simply join the data and then map it out

### 4.2.1 Join the data 

```{r}
popdata2023 <- popdata2023 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper))

#Step 5: Merge the spatial data with the population data
mpsz_pop2020 <- left_join(mpsz19_shp_crs3414, popdata2023,
                          by = c("SUBZONE_N" = "SZ"))
```

### 4.2.2 Plot the map

```{r}
tmap_mode("plot")

tm_shape(mpsz_pop2020)+
  tm_fill("0_to_4", 
          style = "quantile", 
          palette = "Blues",
          title = "0_to_4") +
  tm_layout(main.title = "Distribution of 0_to_4 by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```
