---
title: "Spatial Point Patterns Analysis"
author: "Jiale SO"
date: "August 30, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  warning: false
  freeze: true
---

# 1.0 Context:

------------------------------------------------------------------------

Spatial Point Pattern Analysis involves evaluating the pattern or distribution of a set of points on a surface. These points can represent the locations of:

-   Events, such as crimes, traffic accidents, or disease outbreaks, or

-   Business services, like coffee shops and fast food outlets, or facilities such as childcare and eldercare centers.

The specific questions we would like to answer are as follows:

-   are the childcare centres in Singapore randomly distributed throughout the country?

-   if the answer is not, then the next logical question is where are the locations with higher concentration of childcare centres?

# 2.0 Downloading the Data sets

------------------------------------------------------------------------

To provide answers to the questions above, three data sets will be used. They are:

-   `CHILDCARE`, a point feature data providing both location and attribute information of childcare centres. It was downloaded from Data.gov.sg and is in geojson format. [Link here](https://data.gov.sg/datasets?query=child&page=1&coverage=&formats=GEOJSON&resultId=d_5d668e3f544335f8028f546827b773b4)

-   `MP14_SUBZONE_WEB_PL`, a polygon feature data providing information of URA 2014 Master Plan Planning Subzone boundary data. It is in ESRI shapefile format. This data set was also downloaded from Data.gov.sg. [Link here](https://data.gov.sg/datasets?query=subzone&page=1&coverage=&resultId=d_d14da225fccf921049ab64238ff473d9)

-   `CostalOutline`, a polygon feature data showing the national boundary of Singapore. It is provided by SLA and is in ESRI shapefile format. Link here

# 3.0 Installing and loading R packages

------------------------------------------------------------------------

In this hands-on exercise, five R packages will be used, they are:

-   [**sf**](https://r-spatial.github.io/sf/), a relatively new R package specially designed to import, manage and process vector-based geospatial data in R.

-   [**spatstat**](https://spatstat.org/), which has a wide range of useful functions for point pattern analysis. In this hands-on exercise, it will be used to perform 1st- and 2nd-order spatial point patterns analysis and derive kernel density estimation (KDE) layer.

-   [**raster**](https://cran.r-project.org/web/packages/raster/) which reads, writes, manipulates, analyses and model of gridded spatial data (i.e. raster). In this hands-on exercise, it will be used to convert image output generate by spatstat into raster format.

-   [**maptools**](https://cran.r-project.org/web/packages/maptools/index.html) which provides a set of tools for manipulating geographic data. In this hands-on exercise, we mainly use it to convert *Spatial* objects into *ppp* format of **spatstat**.

-   [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html) which provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [leaflet](https://leafletjs.com/) API.

Use the code chunk below to install and launch the five R packages.

```{r}
pacman::p_load(sf, raster, spatstat, tmap, tidyverse, rvest, geojsonsf)
```

# 4.0 Spatial Data Wrangling

## 4.1 Importing Spatial Data

::: panel-tabset
### Childcare data

```{r}
childcare_sf <- st_read("data/child-care-services-geojson.geojson") %>%
  st_transform(crs = 3414)

```

### Costal Outline

```{r}
sg_sf <- st_read(dsn = "data", layer="CostalOutline")
```

### mp14_subzone

```{r}
mpsz_sf <- st_read(dsn = "data", layer = "MP14_SUBZONE_WEB_PL")
```
:::

### 4.1.1 DIY:Use the appropriate SF function to retrieve the referencing system information of these geospatial data.

Simple, use the `st_crs` function from SF to check and print the crs information

```{r}
# Retrieve CRS information
childcare_crs <- st_crs(childcare_sf)
sg_crs <- st_crs(sg_sf)
mpsz_crs <- st_crs(mpsz_sf)

# Print CRS information
print(childcare_crs)
print(sg_crs)
print(mpsz_crs)
```

### 4.1.2 DIY: Assign the correct CRS to MPSZ_SF and SG_SF Simple Feature Data frames.

notice that the MPSZ_SF and SG_SF is in `World Geodetic System 1984` format, we need set the correct crs to these data and we can do so using the st transform. We can do so using the transform method

::: panel-tabset
### Changing to CRS

```{r}
mpsz_sf <- st_read(dsn = "data", layer = "MP14_SUBZONE_WEB_PL") %>%
  st_transform(crs = 3414)

sg_sf <- st_read(dsn = "data", layer = "CostalOutline") %>%
  st_transform(crs = 3414)
```

### Check if its the right CRS

```{r}
print(st_crs(mpsz_sf))
print(st_crs(sg_sf))
```
:::

### 4.1.3 Change the referencing System to Singapore National Projected Coordinate System

::: callout-note
**Understanding the CRS in Our Data:**

1.  **MPZ and Coastal Data:**

    -   **CRS**: `SVY21`, which is the Singapore National Projected Coordinate System based on WGS84.

    -   **Description**: This is a common projected coordinate system used in Singapore for accurate mapping.

2.  **Childcare Data:**

    -   **CRS**: `SVY21 / Singapore TM` (Transverse Mercator projection).

    -   **Description**: This is also a projection based on SVY21, specifically using the Transverse Mercator projection. It is very close to the SVY21 system, with minor differences in how the projection is handled.

Given that the map file serves as the base, we want all spatial data to overlay correctly, we should:

1.  **Transform the GeoJSON Data to Match the Map File's CRS**:

    -   Since our MPZ and Coastal data are already in `SVY21` (EPSG:3414), transform the GeoJSON data to `EPSG:3414` as well.

2.  **Rationale**:

    -   This approach ensures that the childcare locations from the GeoJSON data will be accurately plotted within the boundaries and context provided by the map file (MPZ and Coastal data).

    -   It avoids potential issues with misalignment, especially since oour base map data is already set up in a local projection suitable for Singapore.
:::

```{r}
# Transform Childcare data to match the base map's CRS (EPSG:3414)
childcare_sf <- st_read("data/child-care-services-geojson.geojson") %>%
  st_transform(crs = 3414)

# Now, all datasets should be aligned in the same CRS
```

### 4.1.4 Checking for validity of maps

When working with spatial data, it's crucial to ensure that all geometries are valid. Invalid geometries can cause errors in analysis and visualization.

1.  **Checking Validity with `st_is_valid()`**:
2.  **Identifying Invalid Geometries**:
3.  **Fixing Invalid Geometries with `st_make_valid()`**

::: panel-tabset
### MPZ

```{r}
mpsz_validity <- st_is_valid(mpsz_sf)
mpsz_invalid <- which(!mpsz_validity)
if (length(mpsz_invalid) > 0) {
  print("MPZ Invalid!")
  print(mpsz_sf[mpsz_invalid, ])
} else {
  print("it's valid!")
}
```

Notice that MPZ has 9 invalidity of sub zones here, so we have to make it valid through the function make valid. Once it's valid we then check again

```{r}
mpsz_sf <- st_make_valid(mpsz_sf)
mpsz_validity <- st_is_valid(mpsz_sf)
mpsz_invalid <- which(!mpsz_validity)
if (length(mpsz_invalid) > 0) {
  print("MPZ Invalid!")
  print(mpsz_sf[mpsz_invalid, ])
} else {
  print("it's valid!")
}
```

### SG

```{r}
sg_validity <- st_is_valid(sg_sf)
sg_invalid <- which(!sg_validity)
if (length(sg_invalid) > 0) {
  print("SG Invalid!")
  print(mpsz_sf[mpsz_invalid, ])
} else {
  print("it's valid!")
}
```

In SG_SF there's one invalid as well, so we apply the fix.

```{r}
sg_sf <- st_make_valid(sg_sf)
sg_validity <- st_is_valid(sg_sf)
sg_invalid <- which(!sg_validity)
if (length(sg_invalid) > 0) {
  print("SG Invalid!")
  print(mpsz_sf[sg_invalid, ])
} else {
  print("it's valid!")
}
```

### Childcare

Notice that childcare is a geojson data and it houses it's data in the description column, we need to break this up to get more meaningful data.

We can do a simple extraction from the `Description` attribute and map the data better. Assuming that each Table Row (TR) contains a Table Head (TH) and a Table Data (TD), we can map the data accordingly.

```{r}
childcare_validity <- st_is_valid(childcare_sf)
childcare_invalid <- which(!childcare_validity)
if (length(childcare_invalid) > 0) {
  print("ChildCare Invalid!")
  print(childcare_sf[childcare_invalid, ])
} else {
  print("it's valid!")
}


# Ensure the geometry column is preserved
geometry_column <- st_geometry(childcare_sf)
parse_description <- function(html_string) {
  html <- read_html(html_string)
  html <- html %>% html_nodes("tr") %>% .[!grepl("Attributes", .)]
  headers <- html %>% html_nodes("th") %>% html_text(trim = TRUE)
  values <- html %>% html_nodes("td") %>% html_text(trim = TRUE)
  
  # Handle cases where the number of headers and values don't match
  if (length(headers) != length(values)) {
    max_length <- max(length(headers), length(values))
    headers <- c(headers, rep("ExtraHeader", max_length - length(headers)))
    values <- c(values, rep("NULL", max_length - length(values)))
  }
  
  setNames(values, headers)
}

# Apply parsing function, unnest the description fields, and remove the original 'Description' column
childcare_sf <- childcare_sf %>% 
  mutate(Description_parsed = map(Description, parse_description)) %>%
  unnest_wider(Description_parsed) %>%
  select(-Description)  # Remove the original 'Description' column

# Overwrite the 'Name' column with the 'LANDYADDRESSPOINT' column values
childcare_sf <- childcare_sf %>%
  mutate(Name = NAME)  # Overwrite 'Name' with 'LANDYADDRESSPOINT'

# Replace empty strings or NA across all columns with "NULL"
childcare_sf <- childcare_sf %>%
  mutate(across(!geometry, ~ ifelse(is.na(.) | . == "", "NULL", .)))

# Reassign the geometry to the dataframe
st_geometry(childcare_sf) <- geometry_column
# Ensure it's still an sf object
class(childcare_sf)
```
:::

## 4.2 Mapping the geospatial datasets.

::: panel-tabset
### DIY PLOT Map

Using the mapping methods you learned in Hands-on Exercise 3, prepare a static map

```{r}
# Suppress the tmap mode message
suppressMessages({
  tmap_mode("plot")  # Use "view" for an interactive map or "plot" for a static map
})

# Create the map
tm <- tm_shape(mpsz_sf) + 
  tm_polygons(col = "grey", border.col = "black", alpha = 0.5) +  # Base map with subzones
  tm_shape(childcare_sf) + 
  tm_dots(col = "black", size = 0.05) +  # Plot childcare locations as dots
   tm_layout(
    main.title = "Childcare Locations on Singapore Map",
    main.title.position = c("center"),  # Center the title at the top
    outer.margins = c(0.1, 0, 0, 0),  # Adjust outer margins to make space for the title
    legend.outside = TRUE,  # Keep legend outside the map area
    legend.outside.position = "bottom"  # Position the legend at the bottom
  )
tm
```

### DIY View Interactivity

we can also prepare a pin map by using the code chunk below.

```{r}
suppressMessages({
  tmap_mode("view")  # Use "view" for an interactive map or "plot" for a static map
})

tm <- tm_shape(mpsz_sf) + 
  tm_polygons(col = "grey", border.col = "black", alpha = 0.5) +  # Base map with subzones
  tm_shape(childcare_sf) + 
  tm_dots(col = "black", size = 0.05) +  # Plot childcare locations as dots
   tm_layout(
    title = "Childcare Locations on Singapore Map",
    title.position = c("center"),  # Center the title at the top
    outer.margins = c(0.1, 0, 0, 0),  # Adjust outer margins to make space for the title
    legend.outside = TRUE,  # Keep legend outside the map area
    legend.outside.position = "bottom"  # Position the legend at the bottom
  )

tm
```
:::
