---
title: "In Class Exercise 5: Spatio-Temporal Point Patterns Analysis"
author: "Jiale SO"
date: "September 09, 2024"
date-modified: "last-modified"
execute: 
  eval: true  
  echo: true
  freeze: true
  warning: false
  message: false
  cache: true
---

# 1.0 Introduction

A spatio-temporal point process (also called space-time or spatial-temporal point process) is a random collection of points, where each point represents the time and location of an event. Examples of events include incidence of disease, sightings or births of a species, or the occurrences of fires, earthquakes, lightning strikes, tsunamis, or volcanic eruptions.

The analysis of spatio-temporal point patterns is becoming increasingly necessary, given the rapid emergence of geographically and temporally indexed data in a wide range of fields. Several spatio-temporal point patterns analysis methods have been introduced and implemented in R in the last ten years. This chapter shows how various R packages can be combined to run a set of spatio-temporal point pattern analyses in a guided and intuitive way. A real world forest fire events in Kepulauan Bangka Belitung, Indonesia from 1st January 2023 to 31st December 2023 is used to illustrate the methods, procedures and interpretations.

## 1.1 Research Questions

The specific question we would like to answer is:

-   are the locations of forest fire in Kepulauan Bangka Belitung spatial and spatio-temporally independent?

-   if the answer is NO, where and when the observed forest fire locations tend to cluster?

# 2.0 Setup and Requirements

For the purpose of this study, five R packages will be used. They are:

-   `rgdal` for importing geospatial data in GIS file format such as shapefile into R and save them as Spatial\*DataFrame,

-   `maptools` for converting Spatial\* object into ppp object,

-   `raster` for handling raster data in R,

-   `sparr` provides function to estimate fixed and adaptive kernel-smoothed spatial relative risk surfaces via the density-ratio method and perform subsequent inference. fixed-bandwidth spati=temporal density and relative risk estimation is also supported

-   `spatstat` for performing Spatial Point Patterns Analysis such as kcross, Lcross, etc., and

-   `tmap` for producing cartographic quality thematic maps

```{r}
pacman::p_load(sf, raster, spatstat, sparr, tmap, tidyverse, stpp, skimr, GWmodel)
```

Two data sets will be used in this hands-on exercise, they are:

-   Hunan county boundary layer. This is a geospatial data set in ESRI shapefile format.

-   Hunan_2012.csv: This csv file contains selected Hunan’s local development indicators in 2012.

::: panel-tabset
## Importing Hunan ShapeFile

```{r, eval = FALSE}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
hunan <- st_transform(hunan, 32647)

```

## importing Hunan Table

```{r, eval = FALSE}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## Joining

```{r, eval = FALSE}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:3,7,9,16,31,32)
```
:::

```{r, eval = FALSE}
write_rds(hunan, "data/rds/hunan_rds")
```

```{r, echo = FALSE}
hunan_sf <- read_rds("data/rds/hunan_rds")
hunan_sp <- hunan_sf %>%
  as_Spatial()

```

# 3.0 Geographically weighted summary statistic with adaptive bandwidth

## 3.1 Determine adaptive bandwidth

::: panel-tabset
## Cross-Validation

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
            data = hunan_sp,
            approach = "CV f",
            adaptive = TRUE,
            kernel = 'bisquare',
            longlat = T
                 )
```

## AIC

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
            data = hunan_sp,
            approach = "AIC",
            adaptive = TRUE,
            kernel = 'bisquare',
            longlat = T
                 )
```
:::

## 3.2 Fixed Distance

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
            data = hunan_sp,
            approach = "CV f",
            adaptive = FALSE,
            kernel = 'bisquare',
            longlat = T
                 )
```

Computing Geographically weighted summary statistics

```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare",
               adaptive = TRUE,
               longlat = T
               )

```
